Generalized Universe Hierarchies and First-Class Universe Levels

### 2021-02-23 paper.tex
correction

\subsection{Morphisms, Liftings and Inclusions of Families}
\label{sec:morphisms}

@@

Family morphisms are restrictions of so-called \emph{weak morphisms}
\cite{dependentrightadjoints} (or \emph{pseudomorphisms}
\cite{kaposi2019gluing}) of cwfs: a weak morphism which has the identity action
on the base category is exactly a family morphism. A plain family morphism is
still not too interesting, so we add more features.

\begin{mydefinition}
A \emph{lifting} is a family morphism which is bijective on terms. Note that
we can drop the $\ext$-preservation condition from the specification of
liftings, since it follows from the invertible action on terms.
\end{mydefinition}

\begin{notation}
We write $\Lift : \Ty_0\,\Gamma \to \Ty_1\,\Gamma$ for the action of some lifting
on types, $\up : \Tm_0\,\Gamma\,A \to \Tm_1\,\Gamma\,(\Lift\,A)$ for the action
on terms, and $\down$ for the inverse action on terms.
\end{notation}

We may think about the relation between \emph{modalities} and liftings.  The
main difference is that we have no structural restrictions on variables and
contexts. More concretely, every $\Lift$ is dependent right adjoint
\cite{dependentrightadjoints} to the identity functor on the base category, as
we have $\Tm\,(\msf{Id}\,\Gamma)\,A \simeq \Tm\,\Gamma\,(\Lift\,A)$. Hence,
every lifting can be viewed as a degenerate modality.

\paragraph{Two-level type theory}
Assume family structures ($\Ty_0$, $\Tm_0$) and ($\Ty_1$, $\Tm_1$) and a
lifting between them. This corresponds to a basic version of \emph{two-level
  type theory} \cite{twolevel}. This theory can be interpreted as having ($\Ty_1$,
$\Tm_1$) as a metaprogramming layer which can generate object-level programs in
the ($\Ty_0$, $\Tm_0$) layer.

@@

\begin{mydefinition}\label{def:inclusion}
A \emph{family inclusion} is a lifting which preserves all type and term
formers. This assumes that every type former which is contained in the source
family, is also contained in the target family.
\end{mydefinition}

@@

%% \begin{mylemma}
%% Every well-founded relation is irreflexive \cite[Lemma 10.4.2]{hottbook}.
%% \end{mylemma}

@@

Our choice of level structures and diagrams is motivated by the following
reasons. First, we do not need identity morphisms in levels, because they would
be mapped to trivial liftings, which are not interesting in our setting. Second,
we do not need proof-relevant level morphisms, since any parallel pair of family
liftings gives rise to isomorphic types. Concretely, given $p : i < j$ and $q :
i < j$ such that $p \neq q$, we have $\Tm_j\,\Gamma\,(\Lift\,p\,A) \simeq
\Tm_i\,\Gamma\,A \simeq \Tm_j\,\Gamma\,(\Lift\,q\,A)$, and since $\Lift\,p\,A$
and $\Lift\,q\,A$ are in the same family, we can internally prove them
isomorphic using function types and identity types. That said, every construction
in this paper would still work with inverse categories as level structures.

>>>>

\subsection{Morphisms and Inclusions of Families}
\label{sec:morphisms}

@@

Family morphisms are restrictions of so-called \emph{weak morphisms}
\cite{dependentrightadjoints} (or \emph{pseudomorphisms}
\cite{kaposi2019gluing}) of cwfs: a weak morphism which has the identity action
on the base category is exactly a family morphism.

\begin{mylemma} Every family morphism has invertible action on terms, i.e.\ there
is an $F^{-1} : \Tm\,\Gamma\,(F\,A) \to \Tm\,\Gamma\,A$. \emph{Proof:} from the
preservation of context extension by $F$, we know that $F$ is invertible on the
zero De Bruijn index, so for $\q : \Tm\,(\Gamma \ext F\,A)\,(F (A[\p]))$, we
have $F^{-1}\q : \Tm\,(\Gamma \ext F\,A)\,(A[\p])$, such that $F (F^{-1}\,\q) =
\q$. If we have any $t : \Tm\,\Gamma\,(F\,A)$, we also have $(F^{-1}\q)[\id,\,t]
: \Tm\,\Gamma\,A$, and this yields the inverse action $F^{-1}$ which acts on all
terms. \qed
\end{mylemma}

\begin{notation}
In the following, we will write $\Lift : \Ty_0\,\Gamma \to \Ty_1\,\Gamma$ for
the action of some morphism on types, $\up : \Tm_0\,\Gamma\,A \to
\Tm_1\,\Gamma\,(\Lift\,A)$ for the action on terms, and $\down$ for the inverse
action on terms. We will also call the action on types \emph{type lifting}
and the action on terms \emph{term lifting}.
\end{notation}

We may think about the relation between \emph{modalities} and morphisms. The
main difference is that we have no structural restrictions on variables and
contexts. More concretely, every $\Lift$ is dependent right adjoint
\cite{dependentrightadjoints} to the identity functor on the base category, as
we have $\Tm\,(\msf{Id}\,\Gamma)\,A \simeq \Tm\,\Gamma\,(\Lift\,A)$. Hence,
every morphism can be viewed as a degenerate modality.

\paragraph{Two-level type theory}
Assume family structures ($\Ty_0$, $\Tm_0$) and ($\Ty_1$, $\Tm_1$) and a
morphism between them. This corresponds to a basic version of \emph{two-level
  type theory} \cite{twolevel}. This theory can be interpreted as having ($\Ty_1$,
$\Tm_1$) as a metaprogramming layer which can generate object-level programs in
the ($\Ty_0$, $\Tm_0$) layer.

@@

\begin{mydefinition}\label{def:inclusion}
A \emph{family inclusion} is a family morphism which preserves all type and term
formers. This assumes that every type former which is contained in the source
family, is also contained in the target family.
\end{mydefinition}

@@

@@

Our choice of level structures and diagrams is motivated by the following
reasons. First, we do not need identity morphisms in levels, because they would
be mapped to trivial liftings, which are not interesting in our setting. Second,
we do not need proof-relevant level morphisms, since any parallel pair of
morphisms gives rise to isomorphic types. Concretely, given $p : i < j$ and $q :
i < j$ such that $p \neq q$, we have $\Tm_j\,\Gamma\,(\Lift\,p\,A) \simeq
\Tm_i\,\Gamma\,A \simeq \Tm_j\,\Gamma\,(\Lift\,q\,A)$, and since $\Lift\,p\,A$
and $\Lift\,q\,A$ are in the same family, we can internally prove them
isomorphic using function types and identity types. That said, every
construction in this paper would still work with inverse categories as level
structures.
